---
title : 'W271 Lab3'
output: 
  pdf_document:
    latex_engine: xelatex
fontsize: 11pt
geometry: margin=1in
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
#Load libraries
sh <- suppressPackageStartupMessages
sh(library(Hmisc));sh(library(plm));sh(library(ggplot2));sh(library(gridExtra));sh(library(dplyr));sh(library(fpp3))
```

## Question 1

```{r}
# load data
load(file = "driving.RData")
sum(is.na(data))
table(data$state)
```

There are no missing data and this is a balanced panel, 25 years observations for each state. We will proceed for subsequent EDA.

```{r warnings=FALSE}
# dependent variable, totfatrte
p1<-qplot(data$totfatrte,geom="histogram",binwidth =2,main = "Histogram of Fatalities", xlab = "Fatalities",fill=I("blue"),col=I("red"),alpha=I(.2),xlim=c(0,60))
p2 <- ggplot(data, aes(factor(state), totfatrte))+geom_boxplot(aes(fill = factor(state)), show.legend = FALSE)+ggtitle("Fatalities by State")+theme(plot.title = element_text(lineheight=1)) 
p3 <- ggplot(data, aes(factor(year), totfatrte))+geom_boxplot(aes(fill = factor(year)), show.legend = FALSE)+ggtitle("Fatalities by Year")+theme(plot.title = element_text(lineheight=1)) 
p4<-data %>% group_by(year)%>%summarise(mean_group=mean(totfatrte))%>%ggplot(aes(x=year, y=mean_group)) + geom_point() +geom_line() + ggtitle("Fatalities Mean Plot across Year") + ylab("Mean Fatalities")
grid.arrange(p1,p2,p3,p4,nrow=2)
```

```{r, warning=FALSE}
# seatbelt
p1<-qplot(data$seatbelt,geom="histogram",binwidth =0.2,main = "Histogram of Seatbelt", xlab = "Seatbelt",fill=I("blue"),col=I("red"),alpha=I(.2),xlim=c(-0.5,2.5))
p2<-ggplot(data, aes(factor(state), seatbelt))+geom_jitter()+ggtitle("Seatbelt by State") + theme(plot.title = element_text(lineheight=1)) 
p3<-ggplot(data, aes((year), seatbelt))+geom_jitter() +ggtitle("Seatbelt by Year") + theme(plot.title = element_text(lineheight=1)) 
p4<-data %>% group_by(year)%>%summarise(mean_group=mean(seatbelt))%>%ggplot(aes(x=year, y=mean_group)) + geom_point() +geom_line() + ggtitle("Mean Plot across Year") + ylab("Mean Seatbelt")
grid.arrange(p1,p2,p3,p4,nrow=2)
```

```{r, warning=FALSE}
# bac08
p1<-qplot(data$bac08,geom="histogram",binwidth =0.2,main = "Histogram of bac08", xlab = "bac08",fill=I("blue"),col=I("red"),alpha=I(.2),xlim=c(-0.5,1.5))
p2<-ggplot(data, aes(factor(state), bac08))+geom_jitter()+ggtitle("Blood Alcohol Content 08 by State")+theme(plot.title = element_text(lineheight=1)) 
p3<-ggplot(data, aes((year), bac08))+geom_jitter()+ggtitle("Blood Alcohol Content 08 by Year")+theme(plot.title = element_text(lineheight=1))
p4<-data %>% group_by(year)%>%summarise(mean_group=mean(bac08))%>%ggplot(aes(x=year, y=mean_group)) + geom_point() +geom_line() + ggtitle("Mean Plot across Year") + ylab("Mean Blood Alcohol Content 08")
grid.arrange(p1,p2,p3,p4,nrow=2)
```

```{r warning=FALSE}
# bac10
p1<-qplot(data$bac10,geom="histogram",binwidth =0.2,main = "Histogram of bac10", xlab = "bac10",fill=I("blue"),col=I("red"),alpha=I(.2),xlim=c(-0.5,1.5))
p2<-ggplot(data, aes(factor(state), bac10))+geom_jitter()+ggtitle("Blood Alcohol Content 10 by State")+theme(plot.title = element_text(lineheight=1)) 
p3<-ggplot(data, aes((year), bac10))+geom_jitter()+ggtitle("Blood Alcohol Content 10 by Year")+theme(plot.title = element_text(lineheight=1))
p4<-data %>% group_by(year)%>%summarise(mean_group=mean(bac10))%>%ggplot(aes(x=year, y=mean_group)) + geom_point() +geom_line() + ggtitle("Mean Plot across Year") + ylab("bac10")
grid.arrange(p1,p2,p3,p4,nrow=2)
```

```{r, warnings=FALSE}
# perse
p1<-qplot(data$perse,geom="histogram",binwidth =0.2,main = "Histogram of perse", xlab = "perse",fill=I("blue"),col=I("red"),alpha=I(.2),xlim=c(-0.5,1.5))
p2<-ggplot(data, aes(factor(state), perse))+geom_jitter()+ggtitle("License Revocation by State")+theme(plot.title = element_text(lineheight=1)) 
p3<-ggplot(data, aes((year), perse))+geom_jitter()+ggtitle("License Revocation by Year")+theme(plot.title = element_text(lineheight=1)) 
p4<-data %>% group_by(year)%>%summarise(mean_group=mean(perse))%>%ggplot(aes(x=year, y=mean_group)) + geom_point() +geom_line() + ggtitle("Mean Plot across Year") + ylab("perse")
grid.arrange(p1,p2,p3,p4,nrow=2)
```

```{r warning=FALSE}
# sl70plus
p1<-qplot(data$perse,geom="histogram",binwidth =0.2,main = "Histogram of sl70plus", xlab = "perse",fill=I("blue"),col=I("red"),alpha=I(.2),xlim=c(-0.5,1.5))
p2<-ggplot(data, aes(factor(state), sl70plus))+geom_jitter()+ggtitle("Speed Limit 70 plus by State")+theme(plot.title = element_text(lineheight=1)) 
p3<-ggplot(data, aes((year), sl70plus))+geom_jitter()+ggtitle("Speed Limit 70 plus by Year")+theme(plot.title = element_text(lineheight=1)) 
p4<-data %>% group_by(year)%>%summarise(mean_group=mean(sl70plus))%>%ggplot(aes(x=year, y=mean_group)) + geom_point() +geom_line() + ggtitle("Mean Plot across Year") + ylab("sl70plus")
grid.arrange(p1,p2,p3,p4,nrow=2)
```

```{r warning=FALSE}
# gdl
p1<-qplot(data$perse,geom="histogram",binwidth =0.2,main = "Histogram of gdl", xlab = "gdl",fill=I("blue"),col=I("red"),alpha=I(.2),xlim=c(-0.5,1.5))
p2<-ggplot(data, aes(factor(state), gdl))+geom_jitter()+ggtitle("Grad. Driver Law by State")+theme(plot.title = element_text(lineheight=1)) 
p3<-ggplot(data, aes((year), gdl))+geom_jitter()+ggtitle("Grad. Driver Law by Year")+theme(plot.title = element_text(lineheight=1)) 
p4<-data %>% group_by(year)%>%summarise(mean_group=mean(gdl))%>%ggplot(aes(x=year, y=mean_group)) + geom_point() +geom_line() + ggtitle("Mean Plot across Year") + ylab("gdl")
grid.arrange(p1,p2,p3,p4,nrow=2)
```

```{r warning=FALSE}
p1<-qplot(data$perc14_24,geom="histogram",binwidth = 0.2,main = "Histogram of % population 14-24",xlab = "percentage",fill=I("blue"),col=I("red"),alpha=I(.2),xlim=c(11,21))
p2<-ggplot(data, aes(factor(state), perc14_24))+geom_boxplot(aes(fill = factor(state)),show.legend=FALSE)+ggtitle("% population 14-24 by State")+theme(plot.title = element_text(lineheight=1)) 
p3<-ggplot(data, aes((year), perc14_24))+geom_jitter()+ggtitle("% population 24-24 by Year")+theme(plot.title = element_text(lineheight=1)) 
p4<-data %>% group_by(year)%>%summarise(mean_group=mean(perc14_24))%>%ggplot(aes(x=year, y=mean_group)) + geom_point() +geom_line() + ggtitle("Mean Plot across Year") + ylab("Mean % population 14-24")
grid.arrange(p1,p2,p3,p4,nrow=2)
```

```{r warning=FALSE}
p1<-qplot(data$unem,geom="histogram",binwidth = 0.5,main="Histogram of Unemployment",xlab = "perc14_24",fill=I("blue"),col=I("red"),alpha=I(.2),xlim=c(0,20))
p2<-ggplot(data, aes(factor(state), unem))+geom_boxplot(aes(fill = factor(state)),show.legend=FALSE)+ggtitle("Unemployment by State")+theme(plot.title = element_text(lineheight=1)) 
p3<-ggplot(data, aes((year), unem))+geom_jitter()+ggtitle("Unemployment by Year")+theme(plot.title = element_text(lineheight=1)) 
p4<-data %>% group_by(year)%>%summarise(mean_group=mean(unem))%>%ggplot(aes(x=year, y=mean_group)) + geom_point() +geom_line() + ggtitle("Mean Plot across Year") + ylab("Mean Unemployment")
grid.arrange(p1,p2,p3,p4,nrow=2)
```

```{r warning=FALSE}
# vehicmilespc
p1<-qplot(data$vehicmilespc,geom="histogram",binwidth = 200,main = "Histogram of Vehicle Miles Traveled PC",xlab = "miles (B)",fill=I("blue"),col=I("red"),alpha=I(.2),xlim=c(4000,19000))
p2<-ggplot(data, aes(factor(state),vehicmilespc))+geom_boxplot(aes(fill=factor(state)),show.legend=FALSE)+ggtitle("Vehicle Miles Traveled PC by State")+theme(plot.title = element_text(lineheight=1)) 
p3<-ggplot(data, aes((year), vehicmilespc))+geom_smooth(method='gam',formula=y~s(x,bs="cs"))+geom_jitter()+ggtitle("Vehicle Miles Traveled PC by Year")+theme(plot.title = element_text(lineheight=1)) 
p4<-data %>% group_by(year)%>%summarise(mean_group=mean(vehicmilespc))%>%ggplot(aes(x=year, y=mean_group)) + geom_point() +geom_line() + ggtitle("Mean Plot across Year") + ylab("Mean Vehicle Miles")
grid.arrange(p1,p2,p3,p4,nrow=2)
```

## Question 2

Variable _totfatrte_ is defined as total number of fatalities in 100,000 population.
```{r}
byYear.mean <- aggregate(data, by=list(data$year), FUN=mean)
mean.totfatrte.df = round(data.frame(year=1980:2004, mean.totfatrte=byYear.mean$totfatrte), 2)
t(mean.totfatrte.df)
as_tsibble(mean.totfatrte.df,index=year)%>%autoplot(mean.totfatrte)+ggtitle("Mean Fatalities by Year")+ylab("Fatalities")
```

Mean of total fatalities show decreasing trend over years. After year 1992, when mean fatalities drop below 18, this number show a stable trend.

```{r}
# Linear Regression
fit.lm <- lm(totfatrte ~ factor(year), data=data)
summary(fit.lm)
```

F-statistic is 7.164 with p-value significantly below threshold level. Using year as explanatory is significant at 95% level. This show that total fatalities is decreasing over time and it is statistically significant. Driving became safer over time.

## Question 3

Variables bac08, bac10, perse, sbprim, sbsecon, sl70plus, gdl are supposed to be binary variables. But due to the fact that some states implemented the law in middle of year, some of the these variables have values between 0 and 1. For correct modeling of binary variables, we need all values to be 0 or 1, for approximation, we will round the values to be 0 or 1.

```{r}
data.round <- data;
data.round$bac08<-factor(round(data$bac08), levels=c(0,1))
data.round$bac10<-factor(round(data$bac10), levels=c(0,1))
data.round$perse<-factor(round(data$perse), levels=c(0,1))
data.round$sbprim<-factor(round(data$sbprim), levels=c(0,1))
data.round$sbsecon<-factor(round(data$sbsecon), levels=c(0,1))
data.round$sl70plus<-factor(round(data$sl70plus), levels=c(0,1))
data.round$gdl<-factor(round(data$gdl), levels=c(0,1))
fit.lm2 <- lm(totfatrte ~ factor(year)+bac08+bac10+perse+sbprim+sbsecon+sl70plus+gdl+perc14_24+unem+vehicmilespc, data=data.round)
summary(fit.lm2)
par(mfrow=c(2,2));plot(fit.lm2);
par(mfrow=c(1,1));acf(fit.lm2$residuals, main="ACF of Residuals");
```

Variables _bac08_ and _bac10_ are binary indicator variables, indicating if a state had law of blood alcohol content of level 0.08% and 0.10% repectively. From mean plot of variables _bac08_ and _bac10_ in EDA, we see that majority of state start with no law on blood alcohol content, and then implementing a 0.10% limit, and then a more strict limit of 0.08%. Coefficient of _bac10_ can be interpreted as, states with blood alcohol content limit 0.10% law have 1.238 less fatalities per 100,000 population.Coefficient of _bac08_ can be interpreted as, states with blood alcohol content limit 0.08% law have 2.194 less fatalities per 100,000 population. 

Variable _perse_ (per se law) has p-value of 0.027433 in pooled OLS result. This variable is statistically significant at 95% level. It shows that there is empirical evidence that per se law has impact on fatalities.

Variable _sbprim_ (primary seat belt law) has p-value of 0.847868 in pooled OLS result. This variable is not statistically significant at 95% level. It shows that there is not empirical evidence that primary seat belt law has impact on fatalities.

One thing to note is that, from regression diagnostic, we observed heteroskedasticity on residuals from scale-location plot and serial correlations on residuals from ACF graph. Serial correlations on residuals suggest there is unobserved fixed effects. Serial correlations and heteroskedasticity on residuals suggest the test statistics in pooled OLS result are not valid.

## Question 4
```{r}
data.panel = pdata.frame(data.round, index=c("state", "year"))
fit.plm.fe <-plm(totfatrte~bac08+bac10+perse+sbprim+sbsecon+sl70plus+gdl+perc14_24+unem+vehicmilespc,
                 data=data.panel, model='within')
summary(fit.plm.fe)
```

In fixed effect model, the coefficient of _bac10_ is similar to pooled OLS and and the coefficient of _bac08_ is smaller in absolute value. _perse_ is highly statistically significant in fixed effects model but it was marginally statistically significant in pooled OLS. _sbprim_ is highly statistically significant in fixed effects model but it was not statistically significant in pooled OLS.

Result from fixed effect model is more reliable. In pooled OLS, we have to assume no unobserved fixed effects, otherwise test statistics are not valid. While in fixed effects model, we are allowed to have unobserved fixed effects present in population model and this fixed effect is allowed to be correlated with explanatory variables. In ACF graph of pooled OLS residuals, we see that serial correlations and this suggests the present of unobserved effect. Therefore assumptions of OLS are not met and pooled OLS result is not reliable. Fixed effect model is the preferred choice.

## Question 5
```{r}
fit.plm.re <- plm(totfatrte~bac08+bac10+perse+sbprim+sbsecon+sl70plus+gdl+perc14_24+unem+vehicmilespc,
                 data=data.panel, model='random')
phtest(fit.plm.fe, fit.plm.re)
```

P-value is smaller than 0.05, we can reject null hypothesis that random effect model is preferred. Fixed Effect model should be chosen for our analysis.

## Question 6
Increase miles driven per capita by 1000, the expect total fatalities per 100,000 population increase by 0.00024665 * 1000 = 0.24665, holding all other variables constant.

## Question 7
Estimators are not efficient. All statistical inference are not valid. If unobserved effect is uncorrelated with all explanatory variables, estimators are consistent, otherwise estimators are not consistent.

